---
title: "Understanding the Primary Socioeconomic Drivers of State Test Scores, Grades 3-8"
author: "Mohammed Abdalla, Miriam Luka, Jayant Kumar, Kashif Awan"
output: pdf_document
---

```{r setup, include=FALSE}
# echo = FALSE hides the code itself; results, tables and figures still print.
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

# This .Rmd lives in notebooks/, but all data paths below are written relative

knitr::opts_knit$set(root.dir = rprojroot::find_root(rprojroot::is_git_root))
```


# Introduction

Standardized test scores are one of the statistics available for nearly every public school district in the country, making them a common measure to assess whether our education system is serving students well. Recent reporting on a generation-long decline in reading scores, paired with proposed cuts to federal Department of Education funding, has drawn increased attention to the matter of where funding should be allocated to improve test scores and more generally educational outcomes. Before discussing funding allocation, policymakers require a clear understanding of the data itself and the potential shared characteristics among school districts with lower test scores.

Socioeconomic status is a characteristic consistently mentioned in this conversation. Built off the correlations indicating that districts with higher incidents of poverty, and less educational attainment among adults tend to have lower-performing school test scores. But socioeconomic status bundles together several distinct conditions: income, education, employment, family structure, and use of public assistance (such as SNAP benefits utilization, etc.). Understanding and quantifying the relationship between socioeconomic status indicators and district state test scores with increased specificity is crucial when attempting to improve outcomes. A policymaker deciding how to allocate funding should understand not just that socioeconomic status is important, but which dimensions of it are most strongly associated with test performance, and how large those associations are in practice.

We approach this question as data consultants for a coalition of school boards seeking to better understand the primary drivers of achievement gaps in their districts. Our research question is purely descriptive: what is the relationship between district-level socioeconomic characteristics and average student test performance across U.S. school districts? We do not claim that any of these characteristics cause test scores to rise or fall, only that describing the strength and shape of these relationships is a necessary first step before anyone can design a targeted, evidence-based funding policy or a causal study of one.

# Data

## Source and Provenance

We use the Stanford Education Data Archive (SEDA), version 6.0 (Reardon et al., Stanford University), a publicly available dataset built from over 480 million standardized test records submitted by U.S. states under federal reporting requirements. SEDA converts each state's own assessments onto a common scale by linking them to the National Assessment of Educational Progress (NAEP), producing test scores that are comparable across states and years. We combine two SEDA files, both at the school-district level:

- **`seda_geodist_annual_cs_6.0`** (achievement): average math and reading/language-arts performance for grades 3–8,

- **`seda_cov_geodist_annual_6.0`** (covariates): district-level socioeconomic and demographic characteristics compiled from the American Community Survey and other federal sources.

For additional information regarding the methodology the SEDA utilized in collecting this data, as well as the files themselves, please visit their website at: https://edopportunity.org/opportunity/data/downloads/#documentation-6

## Data Wrangling

From the achievement file, we keep only the `all students` aggregate (filtering on col `subcat` = `all`), which reports the achievement data in aggregate instead of by subgroups such as race or gender. Furthermore for both our achievement file and covariate file we will filter for `year` = **2018**, as this is the most recent pre-pandemic year available to completely avoid any issues that may have resulted from COVID. Resulting in each row representing one achievement observation per district for 2018. We then join our data to the 2018 covariate file using the district identifier `sedalea`, which is common to both files. Districts that appear in only one file are dropped via an inner join, and rows missing any of our six predictor variables are also dropped. 

Before dropping missing values, the 2018 achievement file has 11,139 district-level **all students** observations and the 2018 covariate file has 12,690. The inner join drops **1,551 covariate districts (12.2%)** that have no 2018 achievement estimate in any subgroup, which is SEDA's suppression rule for districts with too few tested students. The loss runs one way only: every district with an achievement estimate has a matching covariate record, so the join itself discards nothing from the achievement side. A further **95 matched districts** are missing one or more of our six socioeconomic predictors and are dropped as well, leaving a population dataset of **11,044** observations, one per school district. (For reference, 324 districts in the covariate file carry no socioeconomic data at all; 95 of those are among the 11,139 that matched, and are the ones removed here.) As discussed in our previous proposal, districts are not drawn independently, districts neighboring each other likely have similar labor markets, demographics, and state funding resources which would naturally result in iid being violated. We will address this directly with a random sampling method rather than fitting on the full population with a resulting sample size of 1,000.

# Data Prep

```{r dataprep, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
library(readr)
library(tidyr)

# Load raw SEDA files (checked into the repo, already subset to 2018)
seda_achieve <- read_csv("data/raw/seda_geodist_2018.csv",
  show_col_types = FALSE
)
seda_cov <- read_csv("data/raw/seda_cov_geodist_2018.csv",
  show_col_types = FALSE
)

# Achievement file: district-level "all students" estimate, 2018 only
achieve_2018 <- seda_achieve %>%
  filter(year == 2018, subcat == "all", subgroup == "all", gap == 0)

# Covariate file: 2018 socioeconomic predictors
cov_2018 <- seda_cov %>%
  filter(year == 2018) %>%
  select(
    sedalea, lninc50all, baplusall, povertyall,
    unempall, snapall, single_momall, totenrl, perwht,
    perblk, perhsp, perasn, pernam
  )

# Join achievement and covariates on district id (sedalea)
seda_population_file_2018 <- achieve_2018 %>%
  inner_join(cov_2018, by = "sedalea") %>%
  drop_na()

nrow(seda_population_file_2018)
```


# Sampling Methodology

We dropped districts missing any analysis variable, leaving a clean population frame. From it we drew a simple random sample of 1,000 districts, every district had equal probability of selection and each draw was independent, satisfying IID by design. We fixed the random seed for reproducibility.
Before any analysis, we split the sample into 300 exploration districts and 700 confirmation districts. All modeling decisions were made on the exploration set only; the confirmation set was not touched until final estimation. Five automated checks verified the sample was correctly constructed before proceeding.
``` {r sampling, message=FALSE, warning=FALSE, results='hide' }
library(tidyverse)
set.seed(42) # answer of the universe
TARGET_N <- 1000
EXPLORE_FRAC <- 0.30
# (subgroup == "all", gap == FALSE, year 2018).
pop <- seda_population_file_2018
id_vars <- c("sedalea", "sedaleaname", "stateabb", "fips")
outcome <- "cs_mn_avg_ol"
outcome_se <- "cs_mn_avg_ol_se"
ses_comp <- c(
  "baplusall",
  "povertyall",
  "unempall",
  "snapall",
  "single_momall",
  "lninc50all"
)
enrollment <- "totenrl"
race <- c("perwht", "perblk", "perhsp", "perasn", "pernam")
analysis_vars <- c(outcome, ses_comp, enrollment, race)
frame <- pop %>%
  select(all_of(c(id_vars, outcome_se, analysis_vars)))
n_start <- nrow(frame)
frame_clean <- frame %>%
  filter(totenrl > 0) %>%
  drop_na(all_of(analysis_vars))

n_dropped <- n_start - nrow(frame_clean)
cat(sprintf(
  "Frame: %d districts -> %d after cleaning (dropped %d, %.1f%%)\n",
  n_start, nrow(frame_clean), n_dropped, 100 * n_dropped / n_start
))
# validate its IID
stopifnot(TARGET_N <= nrow(frame_clean))
sample_df <- frame_clean %>%
  slice_sample(n = TARGET_N) %>%
  mutate(split = if_else(row_number() <= floor(EXPLORE_FRAC * TARGET_N),
    "exploration", "confirmation"
  ))
run_checks <- function(df) {
  results <- c()
  stopifnot(nrow(df) == TARGET_N)
  results <- c(results, sprintf("PASS  Test 1: sample has %d rows", nrow(df)))
  stopifnot(!any(is.na(df %>% select(all_of(analysis_vars)))))
  results <- c(results, "PASS  Test 2: no NAs in the analysis variables")
  stopifnot(all(df$totenrl > 0))
  results <- c(results, "PASS  Test 3: every district has totenrl > 0")
  stopifnot(n_distinct(df$sedalea) == nrow(df))
  results <- c(results, "PASS  Test 4: all sedalea district ids are unique")
  n_expl <- sum(df$split == "exploration")
  n_conf <- sum(df$split == "confirmation")
  stopifnot(n_expl + n_conf == nrow(df))
  results <- c(results, sprintf(
    "PASS  Test 5: split adds up (exploration=%d, confirmation=%d)",
    n_expl, n_conf
  ))
  cat(paste(results, collapse = "\n"), "\n")
  cat(sprintf("\n%d/5 checks passed.\n", length(results)))
}
run_checks(sample_df)
cat(sprintf(
  "Done! All checks successfully pass for our iid sample of size = %d\n",
  TARGET_N
))
```

# Operationalization and data visualisation

```{r jayant-eda-setup, message=FALSE, warning=FALSE}
## PURPOSE: build `d`, the exploration set this section works from, plus the
##          labels, palette and tallies the rest of the section reads.
##          Attaches to objects created by the Data Prep and Sampling chunks
##          above; reads no files of its own.
library(patchwork)
library(moments)
library(car)
library(knitr)

SES_VARS <- c(
  lninc50all    = "Log median household income",
  baplusall     = "Adults with a bachelor's degree or higher (%)",
  povertyall    = "Households below the poverty line (%)",
  unempall      = "Unemployment rate (%)",
  snapall       = "Households receiving SNAP (%)",
  single_momall = "Households headed by a single mother (%)"
)

SEDA_PALETTE <- c(
  City = "#0072B2", Suburb = "#E69F00",
  Town = "#009E73", Rural = "#CC79A7"
)

## ggplot's default grey panel competes with the data for attention; a minimal
## theme keeps the ink on the points and the fitted line.
theme_set(theme_minimal(base_size = 10))

## urbanicity is not a raw column, and the four locale columns are a trap: they
## look like dummies but are the FRACTION of each district's students in that
## locale type, summing to 1. Testing `urban == 1` silently drops thousands of
## districts, so we take the modal locale instead.
locale_cols <- c(City = "urban", Suburb = "suburb", Town = "town", Rural = "rural")

covariates_raw <- seda_cov %>% filter(year == 2018)
achievement <- achieve_2018

urbanicity_lookup <- covariates_raw %>%
  transmute(sedalea,
    urbanicity = names(locale_cols)[
      max.col(across(all_of(unname(locale_cols))), ties.method = "first")
    ]
  )

## THE line that the confirmation-set swap changes, and the only one.
d <- sample_df %>%
  filter(split == "exploration") %>%
  left_join(urbanicity_lookup, by = "sedalea") %>%
  mutate(
    urbanicity  = factor(urbanicity, levels = c("City", "Suburb", "Town", "Rural")),
    log_poverty = log(povertyall)
  )

## Provenance figures, recomputed from the Data Prep chunk so the numbers quoted
## in the Data Wrangling prose cannot drift out of sync with the pipeline.
n_matched <- nrow(inner_join(select(achieve_2018, sedalea),
  select(cov_2018, sedalea),
  by = "sedalea"
))
n_suppressed <- nrow(covariates_raw) - n_matched

## Suppressed-vs-retained comparison, used inline to state the scope condition.
kept_flag <- covariates_raw %>%
  mutate(Status = if_else(sedalea %in% achievement$sedalea, "Retained", "Suppressed"))
scope <- kept_flag %>%
  group_by(Status) %>%
  summarise(
    enrol_med = median(totenrl, na.rm = TRUE),
    enrol_p25 = quantile(totenrl, .25, na.rm = TRUE),
    poverty = 100 * mean(povertyall, na.rm = TRUE),
    baplus = 100 * mean(baplusall, na.rm = TRUE),
    rural = 100 * mean(max.col(across(all_of(unname(locale_cols))),
      ties.method = "first"
    ) == 4),
    .groups = "drop"
  )
sc <- function(status, col) scope[[col]][scope$Status == status]

stopifnot(nrow(d) == 300, !any(is.na(d$urbanicity)), "log_poverty" %in% names(d))
```

```{r form-fits}
## PURPOSE: Fit the competing functional forms once, early, because the figure in
## the visualisation section draws its fitted line from m_log and the
## appendix table compares all three.
m_lin <- lm(cs_mn_avg_ol ~ povertyall, d)
m_quad <- lm(cs_mn_avg_ol ~ povertyall + I(povertyall^2), d)
m_log <- lm(cs_mn_avg_ol ~ log_poverty, d)
```

# Operationalization

Two concepts have to become variables: *student achievement* and *socioeconomic status*. Neither mapping is obvious, and both admit alternatives we rejected for stated reasons.

| | Variable | Column | Definition |
|---|---|---|---|
| **Outcome (Y)** | Test performance | `cs_mn_avg_ol` | District mean score, math and reading, grades 3--8, in national SD units |
| **Predictors (X)** | Income | `lninc50all` | Log of median household income |
| | Educational attainment | `baplusall` | % of adults with a bachelor's degree or higher |
| | Poverty | `povertyall` | % of residents below the poverty line |
| | Unemployment | `unempall` | % unemployment rate |
| | SNAP receipt | `snapall` | % of households receiving SNAP benefits |
| | Single motherhood | `single_momall` | % of households headed by a single mother |

## The outcome: district test performance

SEDA offers two candidate outcome columns. We use `cs_mn_avg_ol` (the ordinary least-squares estimate) rather than `cs_mn_avg_eb` (an empirical-Bayes estimate that shrinks small districts toward the state mean). Shrinkage would improve precision but at the cost of *deliberately* pulling small districts toward a common value exactly the between-district variation we are trying to describe. For a descriptive study the unshrunken estimate is the honest choice.

```{r y-summary,echo=FALSE, results="hide", fig.cap="Distribution of the outcome on the exploration set. The reference column gives the value each statistic would take for a perfectly symmetric normal distribution, so the last two rows can be read without further context."}
## PURPOSE: Describe the outcome: spread, shape, and whether it clumps at either extreme.
y <- d$cs_mn_avg_ol

y_stats <- tibble(
  Statistic = c(
    "Districts", "Distinct values", "Districts within 0.05 of the minimum",
    "Districts within 0.05 of the maximum",
    "Mean", "SD", "Min", "25th pct", "Median",
    "75th pct", "Max", "Skewness", "Kurtosis"
  ),
  Value = c(
    nrow(d), n_distinct(y),
    sum(y < min(y) + 0.05), sum(y > max(y) - 0.05),
    mean(y), sd(y), min(y),
    quantile(y, .25), median(y), quantile(y, .75), max(y),
    skewness(y), kurtosis(y)
  ),
  Reference = c(
    "--",
    sprintf("%d if no ties", nrow(d)),
    "high value = floor effect", "high value = ceiling effect",
    "--", "--", "--", "--", "--", "--", "--",
    "0 if symmetric", "3 if normal"
  )
) %>%
  # Build Value as character up front: the count wants no decimals, the rest
  # want three. Doing this inside if_else() would format the whole numeric
  # column first and print the count as 300.000000000.
  # The first four rows are counts and want no decimals; the rest want three.
  mutate(Value = c(
    format(Value[1:4], big.mark = ","),
    sprintf("%.3f", Value[-(1:4)])
  ))

kable(y_stats, align = "lrl")
```

The table shows the variable needs no transformation. All `r nrow(d)` districts have a distinct score and none pile up at either extreme, so nothing is censored. Skewness `r round(skewness(d$cs_mn_avg_ol), 2)` and kurtosis `r round(kurtosis(d$cs_mn_avg_ol), 2)` sit near their symmetric and normal reference values. Symmetry is not an OLS requirement the large-sample model assumes nothing about the marginal distribution of $Y$ but it does mean nothing suggests the outcome behaves multiplicatively, so **we leave $Y$ untransformed**. The scale is already interpretable: a coefficient reads directly as a change in standard deviations.

One limitation to state plainly. `cs_mn_avg_ol` is an *estimate*, and SEDA publishes a standard error for each district. About `r round(100 * mean(d$cs_mn_avg_ol_se^2) / var(d$cs_mn_avg_ol), 1)`% of the outcome's variance is measurement error rather than genuine difference between districts, and that error is concentrated in the smallest districts (the correlation between a district's standard error and its log enrolment is `r round(cor(d$cs_mn_avg_ol_se, log(d$totenrl)), 2)`).



## Observations removed

The counts and reasons for every cleaning stage are given in the Data Wrangling section above; the drop that matters is the **`r format(n_suppressed, big.mark = ",")` districts SEDA suppressed** for having too few tested students, plus a further 95 missing one or more covariates. Because SEDA still publishes covariates for the suppressed districts, we can check whether that loss is systematic rather than assuming it. It is, but not in the direction one would guess. Suppressed districts are genuinely smaller — median enrolment `r format(round(sc("Suppressed", "enrol_med")), big.mark = ",")` against `r format(round(sc("Retained", "enrol_med")), big.mark = ",")`, and markedly so at the lower quartile — and somewhat more rural (`r sprintf("%.1f%%", sc("Suppressed", "rural"))` against `r sprintf("%.1f%%", sc("Retained", "rural"))`). But they are **not** poorer or less educated: mean poverty is `r sprintf("%.1f%%", sc("Suppressed", "poverty"))` against `r sprintf("%.1f%%", sc("Retained", "poverty"))`, and the bachelor's-or-higher share `r sprintf("%.1f%%", sc("Suppressed", "baplus"))` against `r sprintf("%.1f%%", sc("Retained", "baplus"))`. The scope condition is therefore narrower than a vague caveat would suggest: **our results describe districts large enough for SEDA to publish a score**, and what is missing is very small, somewhat rural districts rather than a systematically disadvantaged tail.

# The relationship between district poverty and test performance

This is the report's central exhibit: the outcome plotted against the single predictor that carries most of the story, with the fitted model's predictions overlaid.

```{r fig-main, fig.cap="District poverty and average test performance, 300 randomly sampled U.S. school districts, 2018. Each point is one district. The solid line is the fitted relationship from a model of test score on log poverty rate; the shaded band is its 95% confidence interval. Districts are coloured by community type. The relationship is steep among low-poverty districts and flattens as poverty rises.", fig.width=6.5, fig.height=4.2}
## PURPOSE: The report's headline figure: outcome against poverty with fitted predictions overlaid.
pred_grid <- tibble(povertyall = seq(min(d$povertyall), max(d$povertyall), length.out = 200)) %>%
  mutate(log_poverty = log(povertyall))
pred <- bind_cols(pred_grid, as_tibble(predict(m_log, pred_grid, interval = "confidence")))

ggplot(d, aes(povertyall, cs_mn_avg_ol)) +
  geom_hline(yintercept = 0, colour = "grey75", linewidth = .4) +
  geom_point(aes(colour = urbanicity), alpha = .7, size = 1.5) +
  geom_ribbon(
    data = pred, aes(y = fit, ymin = lwr, ymax = upr),
    fill = "grey40", alpha = .18
  ) +
  geom_line(data = pred, aes(y = fit), colour = "grey15", linewidth = .9) +
  scale_colour_manual(values = SEDA_PALETTE, name = "Community type") +
  scale_x_continuous(labels = scales::percent, breaks = scales::pretty_breaks(6)) +
  labs(
    x = "Share of district households below the poverty line",
    y = "Average test score\n(standard deviations from the national average)",
    title = "Poorer districts score lower, but the penalty flattens at high poverty",
    subtitle = "Each point is one U.S. public school district; the grey line is the fitted log-poverty model"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

the figure below displays all `r nrow(d)` sampled districts. The relationship is strongly negative but plainly **not a straight line**: the fitted curve falls steeply across the low-poverty range and flattens beyond roughly 20% poverty. Suburban districts cluster at the top left low poverty, high scores while rural districts, the most numerous group at `r sum(d$urbanicity == "Rural")` of `r nrow(d)`, spread across the full range. Community type is shown for context only; it is not a term in any model.

```{r hypothetical-districts}
## PURPOSE: Translate the fitted model into predicted scores for specific poverty
## rates, read back inline below rather than tabulated (page budget).
hyp <- tibble(`Poverty rate` = c(0.05, 0.10, 0.15, 0.20, 0.30)) %>%
  mutate(
    log_poverty = log(`Poverty rate`),
    `Predicted score` = as.numeric(predict(m_log, tibble(log_poverty = log_poverty)))
  )
```

Read as hypothetical districts, the curve says this. A district at 5% poverty is predicted to score `r sprintf("%+.2f", hyp$"Predicted score"[1])` standard deviations; at 15%, `r sprintf("%+.2f", hyp$"Predicted score"[3])`; at 30%, `r sprintf("%+.2f", hyp$"Predicted score"[5])`. The move from 5% to 15% is associated with a fall of about `r abs(round(hyp$"Predicted score"[3] - hyp$"Predicted score"[1], 2))` standard deviations, while the *larger* move from 15% to 30% is associated with only about `r abs(round(hyp$"Predicted score"[5] - hyp$"Predicted score"[3], 2))`. Ten points of poverty are worth roughly twice as much at the bottom of the range as fifteen points are at the top. That asymmetry is the descriptive finding, and it is why the poverty rate enters the model in logs: a straight line would force both moves to cost the same.

# Appendix: specifications tried

```{r poverty-form}
## PURPOSE: Compare linear, quadratic and log forms for the two predictors where curvature is plausible.

bind_rows(
  tibble(
    Variable = "Poverty rate",
    Form = c("Linear", "Quadratic", "Log"),
    `R-squared` = c(
      summary(m_lin)$r.squared, summary(m_quad)$r.squared,
      summary(m_log)$r.squared
    ),
    `Residual SE` = c(sigma(m_lin), sigma(m_quad), sigma(m_log))
  ),
  tibble(
    Variable = "Bachelor's+ share",
    Form = c("Linear", "Quadratic", "Log"),
    `R-squared` = c(
      summary(lm(cs_mn_avg_ol ~ baplusall, d))$r.squared,
      summary(lm(cs_mn_avg_ol ~ baplusall + I(baplusall^2), d))$r.squared,
      summary(lm(cs_mn_avg_ol ~ log(baplusall), d))$r.squared
    ),
    `Residual SE` = c(
      sigma(lm(cs_mn_avg_ol ~ baplusall, d)),
      sigma(lm(cs_mn_avg_ol ~ baplusall + I(baplusall^2), d)),
      sigma(lm(cs_mn_avg_ol ~ log(baplusall), d))
    )
  )
) %>%
  kable(
    digits = 4, align = "llrr",
    caption = "Three functional forms, tested on both variables where curvature was plausible. Higher R-squared and lower residual SE are better."
  )
```

```{r spec-list}
## PURPOSE: The appendix's required list of specifications tried, each with what
## it taught us. Kept as a table so nothing is claimed without a reason.
tribble(
  ~Specification, ~"What we learned",
  "Y on poverty rate, untransformed",
  "Fits, but the residual pattern and the loess curve both show the straight line is wrong at the extremes.",
  "Y on poverty rate plus a squared term",
  "Curvature is real and strongly significant, so the linear form is rejected. But a fitted parabola eventually turns and predicts higher scores at very high poverty.",
  "Y on log(poverty rate)",
  "Best fit of the three, monotone across the whole range, and interpretable as diminishing returns. Adopted.",
  "Y on bachelor's share, log and squared variants",
  "Neither improves on the linear form, so education enters untransformed. Curvature is not a general feature of these predictors.",
  "Y on all six socioeconomic measures",
  "SNAP receipt flips sign and log income collapses. The measures are too collinear to interpret individually.",
  "Y on log(poverty), education, unemployment, single motherhood",
  "Retains the explanatory power of the six-predictor model with variance inflation near or below three. Preferred specification."
) %>%
  kable(align = "ll")
```

The comparison table above is the evidence for rows one to four: for poverty the log form fits best, while for the bachelor's-degree share the linear form already does, so only poverty is transformed.

```{r final-spec}
## PURPOSE: Fit the preferred specification so the required residuals-vs-fitted
## plot below diagnoses the model actually being proposed.
m_final <- lm(cs_mn_avg_ol ~ log_poverty + baplusall + unempall + single_momall,
  data = d
)
```

```{r fig-resid, fig.cap="Residuals against fitted values for the proposed four-variable model. The local fit (orange) is close to flat with no funnelling, consistent with a correctly specified conditional mean and roughly constant error variance.", fig.width=6.5, fig.height=2.8}
## PURPOSE: Plot residuals against fitted values, to check the conditional mean and error variance.
diag_df <- tibble(fitted = fitted(m_final), resid = resid(m_final))

p_rf <- ggplot(diag_df, aes(fitted, resid)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_point(alpha = .4, size = 1, colour = "#0072B2") +
  geom_smooth(method = "loess", se = FALSE, colour = "#D55E00", linewidth = .8, span = .9) +
  labs(x = "Fitted value", y = "Residual", title = "Residuals vs. fitted")

p_sl <- ggplot(diag_df, aes(fitted, sqrt(abs(resid)))) +
  geom_point(alpha = .4, size = 1, colour = "#0072B2") +
  geom_smooth(method = "loess", se = FALSE, colour = "#D55E00", linewidth = .8, span = .9) +
  labs(
    x = "Fitted value", y = expression(sqrt(abs(residual))),
    title = "Scale--location (variance check)"
  )

p_rf + p_sl
```

the residual plot below is the required residuals-versus-fitted plot for the preferred specification. The local fit is close to flat with no systematic curvature, which is consistent with a correctly specified conditional mean had the poverty rate been left on its raw scale, the curvature documented above would appear here as a parabolic band.


# Model Building

All model choices below are based only on the exploration data. We do not use the confirmation set until the very end, so none of the decisions will introduce bias in the final results.

```{r modeling-setup, message=FALSE, warning=FALSE}
library(lmtest)
library(sandwich)
library(stargazer)

# Filter sample_df down to just the rows tagged "exploration" converts from a tibble to a plain data frame for stargazer
explore <- as.data.frame(sample_df %>% filter(split == "exploration"))
```

We fit three models. **Model 1** regresses test performance on the poverty rate alone, the simplest credible specification and the measure policy discussions lean on most. **Model 2** adds the other five socioeconomic factors, showing how much of poverty's apparent association survives conditioning. **Model 3** adds a squared poverty term to test whether the relationship bends at high poverty. All three report heteroskedasticity-consistent standard errors, since district-level data of this kind commonly has non-constant error variance.

```{r modeling-fit}
# Model 1: starting poing; poverty rate only with no controls
m1 <- lm(cs_mn_avg_ol ~ povertyall, data = explore)

# Model 2: full socioeconomic set that add other factors
m2 <- lm(cs_mn_avg_ol ~ povertyall + lninc50all + baplusall +
  unempall + snapall + single_momall, data = explore)

# Model 3: nonlinearity check adds squared poverty term to test whether the poverty-test score relationship bends at high poverty levels
m3 <- lm(cs_mn_avg_ol ~ povertyall + I(povertyall^2) + lninc50all +
  baplusall + unempall + snapall + single_momall, data = explore)

# District-level socioeconomic data often has non-constant error variance, so we compute HC (heteroskedasticity-consistent)
robust_se <- function(model) sqrt(diag(vcovHC(model)))
```

```{r modeling-table, results='asis', echo=FALSE}
# Set pct to "%" for PDF/LaTeX output, or "%" for HTML output
pct <- if (isTRUE(knitr::is_latex_output())) "\\%" else "%"

# Builds the regression comparision table with robust SEs substituted in for all three
stargazer(m1, m2, m3,
  header = FALSE,
  title = "Regression Results (Exploration Set)",
  se = list(robust_se(m1), robust_se(m2), robust_se(m3)),
  type = if (isTRUE(knitr::is_latex_output())) "latex" else "html",
  dep.var.labels = "Test Performance (SD units from national mean)",
  covariate.labels = c(
    "Poverty Rate", "Poverty Rate Squared", "Log Median Income",
    paste(pct, "Bachelor's Degree+"), "Unemployment Rate",
    paste(pct, "SNAP Recipients"), paste(pct, "Single-Mother Households")
  ),
  digits = 2,
  omit.stat = c("f", "ser"),
  star.cutoffs = c(0.05, 0.01, 0.001),
  font.size = "small",
  no.space = TRUE
)
```


# Assumption Testing
We have compiled several plots to understand the composition of our residuals to assure that our model fits the classic regression model assumptions. From our histogram we can see the errors seem to be be centered around 0, suggestive of a normality of errors. Our qq-residual-plot also confirms this, furthermore using kurtosis() function we can observe a small kurtosis of **1.20956**, indicating a relatively flat distribution with minimal outliers. Regarding Homoskedasticity of errors our scale-location plot seems to suggest this assumption is satisfied as it is mostly flat. Running a BP test to confirm we receive a p-value of **0.3665**, indicating we fail to reject the null assumption of constant variance. Regarding the assumption of Linear Conditional Expectation, checking our fitted-vs-residuals plot we can observe many of our values are close to zero and scattered roughly equally in both a positive and negative direction. With the shape of this plot we can take this assumption as satisfied. Regarding our assumption of no perfect colinnearity our VIF table results are all within 1-6 indicating moderate correlation that should generally be safe for modeling. As discussed in our sampling methodology, we have worked to ensure that our sampling method has satisfied the iid assumption.
```{r Assumption Testing, echo=FALSE, out.width = "70%", results="hide"}
residuals <- residuals(m3)
lmtest::bptest(m3)
kurtosis(residuals) - 3
par(mfrow = c(2, 2))
plot(m3)
```