---
title: "Understanding the Primary Socioeconomic Drivers of State Test Scores, Grades 3-8"
author: "Mohammed Abdalla, Miriam Luka, Jayant Kumar, Kashif Awan"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# This .Rmd lives in notebooks/, but all data paths below are written relative

knitr::opts_knit$set(root.dir = rprojroot::find_root(rprojroot::is_git_root))
```


# Introduction

Standardized test scores are one of the statistics available for nearly every public school district in the country, making them a common measure to assess whether our education system is serving students well. Recent reporting on a generation-long decline in reading scores, paired with proposed cuts to federal Department of Education funding, has drawn increased attention to the matter of where funding should be allocated to improve test scores and more generally educational outcomes. Before discussing funding allocation, policymakers require a clear understanding of the data itself and the potential shared characteristics among school districts with lower test scores.

Socioeconomic status is a characteristic consistently mentioned in this conversation. Built off the correlations indicating that districts with higher incidents of poverty, and less educational attainment among adults tend to have lower-performing school test scores. But socioeconomic status bundles together several distinct conditions: income, education, employment, family structure, and use of public assistance (such as SNAP benefits utilization, etc.). Understanding and quantifying the relationship between socioeconomic status indicators and district state test scores with increased specificity is crucial when attempting to improve outcomes. A policymaker deciding how to allocate funding should understand not just that socioeconomic status is important, but which dimensions of it are most strongly associated with test performance, and how large those associations are in practice.

We approach this question as data consultants for a coalition of school boards seeking to better understand the primary drivers of achievement gaps in their districts. Our research question is purely descriptive: what is the relationship between district-level socioeconomic characteristics and average student test performance across U.S. school districts? We do not claim that any of these characteristics cause test scores to rise or fall, only that describing the strength and shape of these relationships is a necessary first step before anyone can design a targeted, evidence-based funding policy or a causal study of one.

# Data

## Source and Provenance

We use the Stanford Education Data Archive (SEDA), version 6.0 (Reardon et al., Stanford University), a publicly available dataset built from over 480 million standardized test records submitted by U.S. states under federal reporting requirements. SEDA converts each state's own assessments onto a common scale by linking them to the National Assessment of Educational Progress (NAEP), producing test scores that are comparable across states and years. We combine two SEDA files, both at the school-district level:

- **`seda_geodist_annual_cs_6.0`** (achievement): average math and reading/language-arts performance for grades 3–8,

- **`seda_cov_geodist_annual_6.0`** (covariates): district-level socioeconomic and demographic characteristics compiled from the American Community Survey and other federal sources.

For additional information regarding the methodology the SEDA utilized in collecting this data, as well as the files themselves, please visit their website at: https://edopportunity.org/opportunity/data/downloads/#documentation-6

## Data Wrangling

From the achievement file, we keep only the `all students` aggregate (filtering on col `subcat` = `all`), which reports the achievement data in aggregate instead of by subgroups such as race or gender. Furthermore for both our achievement file and covariate file we will filter for `year` = **2018**, as this is the most recent pre-pandemic year available to completely avoid any issues that may have resulted from COVID. Resulting in each row representing one achievement observation per district for 2018. We then join our data to the 2018 covariate file using the district identifier `sedalea`, which is common to both files. Districts that appear in only one file are dropped via an inner join, and rows missing any of our six predictor variables are also dropped. 

Before dropping missing values, the 2018 achievement file has 11,139 district-level **all students** observations and the 2018 covariate file has 12,690. The inner join drops **1,551 covariate districts (12.2%)** that have no 2018 achievement estimate in any subgroup, which is SEDA's suppression rule for districts with too few tested students. The loss runs one way only: every district with an achievement estimate has a matching covariate record, so the join itself discards nothing from the achievement side. A further **95 matched districts** are missing one or more of our six socioeconomic predictors and are dropped as well, leaving a population dataset of **11,044** observations, one per school district. (For reference, 324 districts in the covariate file carry no socioeconomic data at all; 95 of those are among the 11,139 that matched, and are the ones removed here.) As discussed in our previous proposal, districts are not drawn independently, districts neighboring each other likely have similar labor markets, demographics, and state funding resources which would naturally result in iid being violated. We will address this directly with a random sampling method rather than fitting on the full population with a resulting sample size of 1,000.

## Operationalization

**Outcome (Y):** District test performance, `cs_mn_avg_ol` — a standardized effect size measuring how far a district's average test score falls from the national mean, in national standard-deviation units. It is metric and pools math and reading/language-arts performance across grades 3–8.

**Predictors (X), all district-level, metric:**

| Variable | Column | Definition |
|---|---|---|
| Income | `lninc50all` | Log of median household income |
| Educational attainment | `baplusall` | % of adults with a bachelor's degree or higher |
| Poverty | `povertyall` | % of residents below the poverty line |
| Unemployment | `unempall` | % unemployment rate |
| SNAP receipt | `snapall` | % of households receiving SNAP benefits |
| Single motherhood | `single_momall` | % of households headed by a single mother |


# Data Prep

```{r dataprep, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
library(readr)
library(tidyr)

# Load raw SEDA files (checked into the repo, already subset to 2018)
seda_achieve <- read_csv("data/raw/seda_geodist_2018.csv",
  show_col_types = FALSE
)
seda_cov <- read_csv("data/raw/seda_cov_geodist_2018.csv",
  show_col_types = FALSE
)

# Achievement file: district-level "all students" estimate, 2018 only
achieve_2018 <- seda_achieve %>%
  filter(year == 2018, subcat == "all", subgroup == "all", gap == 0)

# Covariate file: 2018 socioeconomic predictors
cov_2018 <- seda_cov %>%
  filter(year == 2018) %>%
  select(
    sedalea, lninc50all, baplusall, povertyall,
    unempall, snapall, single_momall, totenrl, perwht,
    perblk, perhsp, perasn, pernam
  )

# Join achievement and covariates on district id (sedalea)
seda_population_file_2018 <- achieve_2018 %>%
  inner_join(cov_2018, by = "sedalea") %>%
  drop_na()

nrow(seda_population_file_2018)
```


# Sampling Methodology

Data & Sampling

Our data comes from the 2018 wave of the Stanford Education Data Archive (SEDA), a nationally comprehensive source that provides district-level academic performance and demographic information for public school districts across the United States. We work with the all-subgroups population file, which gives us one record per district, and restrict the frame to districts with positive enrollment and complete data on all variables we intend to analyze. That cleaning step leaves us with a well-defined population of districts before we draw a single observation.

The variables we retain reflect the core question. Our outcome is each district's average test score composite. Our socioeconomic predictors — bachelor's degree attainment, poverty rate, unemployment, SNAP receipt, single-mother household share, and log median income — are kept as separate measures rather than collapsed into a single index, precisely because we want to understand which dimensions of socioeconomic status carry the most weight. We also retain total enrollment and race and ethnicity shares (white, Black, Hispanic, Asian, and Native American) as contextual controls.

From that cleaned frame we draw a simple random sample of 1,000 districts without replacement. SRS is the right choice here because it satisfies the IID assumption directly: every district is selected independently and with equal probability, with no clustering or stratification introducing dependence into the sample. A fixed random seed ensures the draw is reproducible.

We then immediately split the sample before conducting any analysis. Three hundred districts (30%) are assigned to an exploration set; the remaining 700 (70%) are held as a confirmation set. All variable selection, distributional checks, and collinearity diagnostics are carried out on the exploration set only. The confirmation set does not inform any modeling decisions — it is reserved entirely for final estimation. This discipline matters because allowing the same data to guide both specification and estimation inflates the risk of finding patterns that do not replicate. Before writing any output, five automated checks verify that the sample is correctly sized, contains no missing values on analysis variables, includes only districts with positive enrollment, has unique district identifiers throughout, and that the exploration and confirmation counts sum correctly to 1,000.

``` {r sampling, message=FALSE, warning=FALSE, results='hide' }
library(tidyverse)
set.seed(42) # answer of the universe
TARGET_N <- 1000
EXPLORE_FRAC <- 0.30
# (subgroup == "all", gap == FALSE, year 2018).
pop <- seda_population_file_2018
id_vars <- c("sedalea", "sedaleaname", "stateabb", "fips")
outcome <- "cs_mn_avg_ol"
outcome_se <- "cs_mn_avg_ol_se"
ses_comp <- c(
  "baplusall",
  "povertyall",
  "unempall",
  "snapall",
  "single_momall",
  "lninc50all"
)
enrollment <- "totenrl"
race <- c("perwht", "perblk", "perhsp", "perasn", "pernam")
analysis_vars <- c(outcome, ses_comp, enrollment, race)
frame <- pop %>%
  select(all_of(c(id_vars, outcome_se, analysis_vars)))
n_start <- nrow(frame)
frame_clean <- frame %>%
  filter(totenrl > 0) %>%
  drop_na(all_of(analysis_vars))

n_dropped <- n_start - nrow(frame_clean)
cat(sprintf(
  "Frame: %d districts -> %d after cleaning (dropped %d, %.1f%%)\n",
  n_start, nrow(frame_clean), n_dropped, 100 * n_dropped / n_start
))
# validate its IID
stopifnot(TARGET_N <= nrow(frame_clean))
sample_df <- frame_clean %>%
  slice_sample(n = TARGET_N) %>%
  mutate(split = if_else(row_number() <= floor(EXPLORE_FRAC * TARGET_N),
    "exploration", "confirmation"
  ))
run_checks <- function(df) {
  results <- c()
  stopifnot(nrow(df) == TARGET_N)
  results <- c(results, sprintf("PASS  Test 1: sample has %d rows", nrow(df)))
  stopifnot(!any(is.na(df %>% select(all_of(analysis_vars)))))
  results <- c(results, "PASS  Test 2: no NAs in the analysis variables")
  stopifnot(all(df$totenrl > 0))
  results <- c(results, "PASS  Test 3: every district has totenrl > 0")
  stopifnot(n_distinct(df$sedalea) == nrow(df))
  results <- c(results, "PASS  Test 4: all sedalea district ids are unique")
  n_expl <- sum(df$split == "exploration")
  n_conf <- sum(df$split == "confirmation")
  stopifnot(n_expl + n_conf == nrow(df))
  results <- c(results, sprintf(
    "PASS  Test 5: split adds up (exploration=%d, confirmation=%d)",
    n_expl, n_conf
  ))
  cat(paste(results, collapse = "\n"), "\n")
  cat(sprintf("\n%d/5 checks passed.\n", length(results)))
}
run_checks(sample_df)
cat(sprintf(
  "Done! All checks successfully pass for our iid sample of size = %d\n",
  TARGET_N
))
```

# Operationalization and data visualisation

```{r jayant-eda-setup, message=FALSE, warning=FALSE}
## PURPOSE: build `d`, the exploration set this section works from, plus the
##          labels, palette and tallies the rest of the section reads.
##          Attaches to objects created by the Data Prep and Sampling chunks
##          above; reads no files of its own.
library(patchwork)
library(moments)
library(car)
library(knitr)

SES_VARS <- c(
  lninc50all    = "Log median household income",
  baplusall     = "Adults with a bachelor's degree or higher (%)",
  povertyall    = "Households below the poverty line (%)",
  unempall      = "Unemployment rate (%)",
  snapall       = "Households receiving SNAP (%)",
  single_momall = "Households headed by a single mother (%)"
)

SEDA_PALETTE <- c(City = "#0072B2", Suburb = "#E69F00",
                  Town = "#009E73", Rural = "#CC79A7")

## ggplot's default grey panel competes with the data for attention; a minimal
## theme keeps the ink on the points and the fitted line.
theme_set(theme_minimal(base_size = 10))

## urbanicity is not a raw column, and the four locale columns are a trap: they
## look like dummies but are the FRACTION of each district's students in that
## locale type, summing to 1. Testing `urban == 1` silently drops thousands of
## districts, so we take the modal locale instead.
locale_cols <- c(City = "urban", Suburb = "suburb", Town = "town", Rural = "rural")

covariates_raw <- seda_cov %>% filter(year == 2018)
achievement    <- achieve_2018

urbanicity_lookup <- covariates_raw %>%
  transmute(sedalea,
            urbanicity = names(locale_cols)[
              max.col(across(all_of(unname(locale_cols))), ties.method = "first")
            ])

## THE line that the confirmation-set swap changes, and the only one.
d <- sample_df %>%
  filter(split == "exploration") %>%
  left_join(urbanicity_lookup, by = "sedalea") %>%
  mutate(
    urbanicity  = factor(urbanicity, levels = c("City", "Suburb", "Town", "Rural")),
    log_poverty = log(povertyall)
  )

locale_share_max <- apply(covariates_raw[unname(locale_cols)], 1, max)

## Provenance tally, rebuilt from the Data Prep and Sampling chunks so the
## dropped-observations table is generated rather than transcribed.
n_matched <- nrow(inner_join(select(achieve_2018, sedalea),
                             select(cov_2018, sedalea), by = "sedalea"))

drop_log <- tibble::tribble(
  ~step, ~n_in, ~n_out, ~reason,
  "Achievement rows to district rows", nrow(seda_achieve), nrow(achieve_2018),
    "Kept subcat/subgroup 'all' and gap == 0. The other rows are subgroup means and subgroup gaps, not district means. A reshape, not a data loss.",
  "Covariate districts to matched districts", nrow(covariates_raw), n_matched,
    "Districts with covariates but no 2018 achievement estimate. SEDA suppresses districts with too few tested students.",
  "Matched districts to complete records", n_matched, nrow(seda_population_file_2018),
    "Missing one or more socioeconomic covariates (listwise deletion).",
  "Frame to analysis sample", nrow(frame_clean), nrow(sample_df),
    "Simple random sample without replacement, set.seed(42).",
  "Analysis sample to exploration set", nrow(sample_df), nrow(d),
    "Confirmation set held out; not opened until final estimation."
)

stopifnot(nrow(d) == 300, !any(is.na(d$urbanicity)), "log_poverty" %in% names(d))
```

```{r form-fits}
## PURPOSE: Fit the competing functional forms once, early, because the figure in
## the visualisation section draws its fitted line from m_log and the
## appendix table compares all three.
m_lin <- lm(cs_mn_avg_ol ~ povertyall, d)
m_quad <- lm(cs_mn_avg_ol ~ povertyall + I(povertyall^2), d)
m_log <- lm(cs_mn_avg_ol ~ log_poverty, d)
```

# Operationalization: evidence for the variable choices

*This section supplies the evidence behind the variable choices summarised in the Operationalization subsection above above; the two should be merged before submission rather than both appearing.* Two concepts have to become variables: *student achievement* and *socioeconomic status*. Neither mapping is obvious, and both admit alternatives we rejected for stated reasons.

## The outcome: district test performance

We operationalise achievement as `cs_mn_avg_ol` a district's mean score across math and reading, grades 3--8, expressed in standard deviations from the national average.

SEDA offers two candidate outcome columns. We use `cs_mn_avg_ol` (the ordinary least-squares estimate) rather than `cs_mn_avg_eb` (an empirical-Bayes estimate that shrinks small districts toward the state mean). Shrinkage would improve precision but at the cost of *deliberately* pulling small districts toward a common value exactly the between-district variation we are trying to describe. For a descriptive study the unshrunken estimate is the honest choice.

```{r y-summary, fig.cap="Distribution of the outcome on the exploration set. The reference column gives the value each statistic would take for a perfectly symmetric normal distribution, so the last two rows can be read without further context."}
## PURPOSE: Describe the outcome: spread, shape, and whether it clumps at either extreme.
y <- d$cs_mn_avg_ol

y_stats <- tibble(
 Statistic = c("Districts", "Distinct values", "Districts within 0.05 of the minimum",
 "Districts within 0.05 of the maximum",
 "Mean", "SD", "Min", "25th pct", "Median",
 "75th pct", "Max", "Skewness", "Kurtosis"),
 Value = c(
 nrow(d), n_distinct(y),
 sum(y < min(y) + 0.05), sum(y > max(y) - 0.05),
 mean(y), sd(y), min(y),
 quantile(y, .25), median(y), quantile(y, .75), max(y),
 skewness(y), kurtosis(y)
 ),
 Reference = c("--",
 sprintf("%d if no ties", nrow(d)),
 "high value = floor effect", "high value = ceiling effect",
 "--", "--", "--", "--", "--", "--", "--",
 "0 if symmetric", "3 if normal")
) %>%
 # Build Value as character up front: the count wants no decimals, the rest
 # want three. Doing this inside if_else() would format the whole numeric
 # column first and print the count as 300.000000000.
 # The first four rows are counts and want no decimals; the rest want three.
 mutate(Value = c(format(Value[1:4], big.mark = ","),
 sprintf("%.3f", Value[-(1:4)])))

kable(y_stats, align = "lrl")
```

The table shows the variable needs no transformation. All `r nrow(d)` districts have a distinct score and none pile up at either extreme, so nothing is censored. Skewness `r round(skewness(d$cs_mn_avg_ol), 2)` and kurtosis `r round(kurtosis(d$cs_mn_avg_ol), 2)` sit near their symmetric and normal reference values. Symmetry is not an OLS requirement the large-sample model assumes nothing about the marginal distribution of $Y$ but it does mean nothing suggests the outcome behaves multiplicatively, so **we leave $Y$ untransformed**. The scale is already interpretable: a coefficient reads directly as a change in standard deviations.

One limitation to state plainly. `cs_mn_avg_ol` is an *estimate*, and SEDA publishes a standard error for each district. About `r round(100 * mean(d$cs_mn_avg_ol_se^2) / var(d$cs_mn_avg_ol), 1)`% of the outcome's variance is measurement error rather than genuine difference between districts, and that error is concentrated in the smallest districts (the correlation between a district's standard error and its log enrolment is `r round(cor(d$cs_mn_avg_ol_se, log(d$totenrl)), 2)`).

## The predictors: socioeconomic status as six measures, not one index

Socioeconomic status could be operationalised as a single composite SEDA ships one, `sesall`. We deliberately do not use it. The research question asks *which dimensions* of socioeconomic status carry the most weight, and a composite answers that question by erasing it. So we keep six separate measures: log median income, adult bachelor's-degree share, poverty rate, unemployment, SNAP receipt, and single-mother household share.

That choice has a cost, and it is worth quantifying rather than glossing.

```{r vif}
## PURPOSE: Quantify how much the predictors inflate each other's variance when entered together.
m_all <- lm(reformulate(names(SES_VARS), "cs_mn_avg_ol"), data = d)
tibble(Variable = SES_VARS[names(vif(m_all))], VIF = as.numeric(vif(m_all))) %>%
 arrange(desc(VIF)) %>%
 kable(digits = 2, caption = "Variance inflation factors, all six predictors entered together.")
```

```{r collinearity-flip, fig.cap="What collinearity does to two of the predictors. Each is shown fitted alone, then fitted alongside the other five. A coefficient that changes sign between the two columns is not measuring a stable relationship."}
## PURPOSE: Show what collinearity does: fit two predictors alone, then alongside the rest.
map_dfr(c("snapall", "lninc50all"), function(v) {
 alone <- lm(reformulate(v, "cs_mn_avg_ol"), data = d)
 tibble(
 Predictor = SES_VARS[v],
 `Slope alone` = coef(alone)[2],
 `p-value alone` = summary(alone)$coefficients[2, 4],
 `Slope with all 6` = coef(m_all)[v],
 `p-value with all 6` = summary(m_all)$coefficients[v, 4]
 )
}) %>%
 kable(digits = 3, align = "lrrrr")
```

The six measures overlap heavily poverty correlates with SNAP receipt at `r round(cor(d$povertyall, d$snapall), 2)` and with log income at `r round(cor(d$povertyall, d$lninc50all), 2)`. The variance inflation factors show the consequence, and the second table makes it concrete: SNAP receipt and log income are both strong, highly significant predictors on their own, but entered alongside the other four SNAP **flips sign** and loses all significance while income collapses toward zero.

Neither variable has stopped mattering; each has been absorbed by its correlates. We therefore **exclude SNAP receipt and log median income** from the preferred specification not because they are unimportant, but because their coefficients in a six-predictor model would not mean what a reader would take them to mean. Poverty is retained over SNAP because it is the more widely understood measure; education and poverty together carry most of what income would have contributed.

## Observations removed

```{r drops}
## PURPOSE: Report the provenance tally: how many districts were removed at each stage, and why.
## Not hand-typed: this is the tally accumulated by log_step() as the cleaning section ran,
## so the table cannot drift out of sync with the pipeline that produced the data.
drop_log %>%
 transmute(
 Step = step,
 `Rows in` = n_in,
 `Rows out` = n_out,
 Dropped = n_in - n_out,
 `%` = round(100 * (n_in - n_out) / n_in, 1),
 Reason = reason
 ) %>%
 kable(caption = "Observations removed at each stage of the pipeline, with reasons. Generated by the cleaning code, not transcribed.")
```

```{r suppressed-n}
## PURPOSE: Extract the count of SEDA-suppressed districts for use in the prose below.
n_suppressed <- drop_log$n_in[2] - drop_log$n_out[2]
```

The first row is a reshape, not a loss: the achievement file stacks subgroup means and achievement gaps alongside district means, and only the district rows are our unit of observation. The drop that matters is the **`r format(n_suppressed, big.mark = ",")` districts SEDA suppressed** for having too few tested students, plus a further 95 missing one or more covariates.

Because SEDA still publishes covariates for the suppressed districts, we can check whether that loss is systematic rather than assuming it.

```{r missingness-check, fig.cap="Suppressed vs. retained districts, compared on the covariates SEDA publishes for both. Enrolment is reported as a median because it is heavily right-skewed; the rest are means."}
## PURPOSE: Compare suppressed districts against retained ones, to test whether the loss is random.
kept_flag <- covariates_raw %>%
 mutate(Status = if_else(sedalea %in% achievement$sedalea,
 "Retained", "Suppressed"))

bind_rows(
 kept_flag %>%
 group_by(Status) %>%
 summarise(
 Measure = "Districts",
 Value = as.character(n()), .groups = "drop"
 ),
 kept_flag %>%
 group_by(Status) %>%
 summarise(
 Measure = "Median enrolment",
 Value = format(round(median(totenrl, na.rm = TRUE)), big.mark = ","),
 .groups = "drop"
 ),
 kept_flag %>%
 group_by(Status) %>%
 summarise(
 Measure = "25th percentile of enrolment",
 Value = format(round(quantile(totenrl, .25, na.rm = TRUE)), big.mark = ","),
 .groups = "drop"
 ),
 kept_flag %>%
 group_by(Status) %>%
 summarise(
 Measure = "Mean poverty rate",
 Value = sprintf("%.1f%%", 100 * mean(povertyall, na.rm = TRUE)),
 .groups = "drop"
 ),
 kept_flag %>%
 group_by(Status) %>%
 summarise(
 Measure = "Mean bachelor's+ share",
 Value = sprintf("%.1f%%", 100 * mean(baplusall, na.rm = TRUE)),
 .groups = "drop"
 ),
 kept_flag %>%
 group_by(Status) %>%
 summarise(
 Measure = "Share rural",
 Value = sprintf("%.1f%%", 100 * mean(
 max.col(across(all_of(unname(locale_cols))), ties.method = "first") == 4
 )),
 .groups = "drop"
 )
) %>%
 pivot_wider(names_from = Status, values_from = Value) %>%
 kable(align = "lrr")
```

This corrects an assumption it would have been easy to make. The suppressed districts are genuinely smaller markedly so at the lower quartile and somewhat more rural. But they are **not** poorer or less educated; on both measures they look slightly better off than the districts retained. The scope condition is therefore narrower than a vague caveat would suggest: **our results describe districts large enough for SEDA to publish a score**, and what is missing is very small, somewhat rural districts rather than a systematically disadvantaged tail.

# The relationship between district poverty and test performance

This is the report's central exhibit: the outcome plotted against the single predictor that carries most of the story, with the fitted model's predictions overlaid.

```{r fig-main, fig.cap="District poverty and average test performance, 300 randomly sampled U.S. school districts, 2018. Each point is one district. The solid line is the fitted relationship from a model of test score on log poverty rate; the shaded band is its 95% confidence interval. Districts are coloured by community type. The relationship is steep among low-poverty districts and flattens as poverty rises.", fig.width=6.5, fig.height=4.2}
## PURPOSE: The report's headline figure: outcome against poverty with fitted predictions overlaid.
pred_grid <- tibble(povertyall = seq(min(d$povertyall), max(d$povertyall), length.out = 200)) %>%
 mutate(log_poverty = log(povertyall))
pred <- bind_cols(pred_grid, as_tibble(predict(m_log, pred_grid, interval = "confidence")))

ggplot(d, aes(povertyall, cs_mn_avg_ol)) +
 geom_hline(yintercept = 0, colour = "grey75", linewidth = .4) +
 geom_point(aes(colour = urbanicity), alpha = .7, size = 1.5) +
 geom_ribbon(data = pred, aes(y = fit, ymin = lwr, ymax = upr),
 fill = "grey40", alpha = .18) +
 geom_line(data = pred, aes(y = fit), colour = "grey15", linewidth = .9) +
 scale_colour_manual(values = SEDA_PALETTE, name = "Community type") +
 scale_x_continuous(labels = scales::percent, breaks = scales::pretty_breaks(6)) +
 labs(
 x = "Share of district households below the poverty line",
 y = "Average test score\n(standard deviations from the national average)",
 title = "Poorer districts score lower, but the penalty flattens at high poverty",
 subtitle = "Each point is one U.S. public school district; the grey line is the fitted log-poverty model"
 ) +
 theme(legend.position = "bottom",
 plot.title = element_text(face = "bold"),
 panel.grid.minor = element_blank())
```

the figure below displays all `r nrow(d)` sampled districts. The relationship is strongly negative but plainly **not a straight line**: the fitted curve falls steeply across the low-poverty range and flattens beyond roughly 20% poverty. Suburban districts cluster at the top left low poverty, high scores while rural districts, the most numerous group at `r sum(d$urbanicity == "Rural")` of `r nrow(d)`, spread across the full range. Community type is shown for context only; it is not a term in any model.

```{r hypothetical-districts, fig.cap="Hypothetical districts: predicted mean test score at selected poverty rates, from the log-poverty model. The final column is the cost of getting to that row from the one above it."}
## PURPOSE: Translate the fitted model into predicted scores for specific poverty rates.
hyp <- tibble(`Poverty rate` = c(0.05, 0.10, 0.15, 0.20, 0.30)) %>%
 mutate(
 log_poverty = log(`Poverty rate`),
 `Predicted score` = as.numeric(predict(m_log, tibble(log_poverty = log_poverty))),
 `Change from row above` = `Predicted score` - lag(`Predicted score`),
 `Poverty rate` = sprintf("%.0f%%", 100 * `Poverty rate`)
 ) %>%
 select(-log_poverty)

hyp %>% kable(digits = 2, align = "lrr")
```

The table translates the curve into districts a reader can picture. Moving a hypothetical district from 5% to 15% poverty is associated with a fall of about `r abs(round(hyp$"Predicted score"[3] - hyp$"Predicted score"[1], 2))` standard deviations; the *larger* move from 15% to 30% is associated with only about `r abs(round(hyp$"Predicted score"[5] - hyp$"Predicted score"[3], 2))`. Ten points of poverty are worth roughly twice as much at the bottom of the range as fifteen points are at the top.

That asymmetry is the descriptive finding, and it is why the poverty rate enters the model in logs. A straight line would force both moves to cost the same and would misrepresent what the data shows.

# Appendix: specifications tried

```{r poverty-form}
## PURPOSE: Compare linear, quadratic and log forms for the two predictors where curvature is plausible.

bind_rows(
 tibble(
 Variable = "Poverty rate",
 Form = c("Linear", "Quadratic", "Log"),
 `R-squared` = c(summary(m_lin)$r.squared, summary(m_quad)$r.squared,
 summary(m_log)$r.squared),
 `Residual SE` = c(sigma(m_lin), sigma(m_quad), sigma(m_log))
 ),
 tibble(
 Variable = "Bachelor's+ share",
 Form = c("Linear", "Quadratic", "Log"),
 `R-squared` = c(
 summary(lm(cs_mn_avg_ol ~ baplusall, d))$r.squared,
 summary(lm(cs_mn_avg_ol ~ baplusall + I(baplusall^2), d))$r.squared,
 summary(lm(cs_mn_avg_ol ~ log(baplusall), d))$r.squared
 ),
 `Residual SE` = c(
 sigma(lm(cs_mn_avg_ol ~ baplusall, d)),
 sigma(lm(cs_mn_avg_ol ~ baplusall + I(baplusall^2), d)),
 sigma(lm(cs_mn_avg_ol ~ log(baplusall), d))
 )
 )
) %>%
 kable(digits = 4, align = "llrr",
 caption = "Three functional forms, tested on both variables where curvature was plausible. Higher R-squared and lower residual SE are better.")
```

```{r spec-list}
## PURPOSE: The appendix's required list of specifications tried, each with what
## it taught us. Kept as a table so nothing is claimed without a reason.
tribble(
 ~Specification, ~"What we learned",
 "Y on poverty rate, untransformed",
 "Fits, but the residual pattern and the loess curve both show the straight line is wrong at the extremes.",
 "Y on poverty rate plus a squared term",
 "Curvature is real and strongly significant, so the linear form is rejected. But a fitted parabola eventually turns and predicts higher scores at very high poverty.",
 "Y on log(poverty rate)",
 "Best fit of the three, monotone across the whole range, and interpretable as diminishing returns. Adopted.",
 "Y on bachelor's share, log and squared variants",
 "Neither improves on the linear form, so education enters untransformed. Curvature is not a general feature of these predictors.",
 "Y on all six socioeconomic measures",
 "SNAP receipt flips sign and log income collapses. The measures are too collinear to interpret individually.",
 "Y on log(poverty), education, unemployment, single motherhood",
 "Retains the explanatory power of the six-predictor model with variance inflation near or below three. Preferred specification."
) %>%
 kable(align = "ll")
```

The comparison table above is the evidence for rows one to four: for poverty the log form fits best, while for the bachelor's-degree share the linear form already does, so only poverty is transformed.

```{r final-spec}
## PURPOSE: Fit the preferred specification so the required residuals-vs-fitted
## plot below diagnoses the model actually being proposed.
m_final <- lm(cs_mn_avg_ol ~ log_poverty + baplusall + unempall + single_momall,
 data = d)
```

```{r fig-resid, fig.cap="Residuals against fitted values for the proposed four-variable model. The local fit (orange) is close to flat with no funnelling, consistent with a correctly specified conditional mean and roughly constant error variance.", fig.width=6.5, fig.height=2.8}
## PURPOSE: Plot residuals against fitted values, to check the conditional mean and error variance.
diag_df <- tibble(fitted = fitted(m_final), resid = resid(m_final))

p_rf <- ggplot(diag_df, aes(fitted, resid)) +
 geom_hline(yintercept = 0, colour = "grey60") +
 geom_point(alpha = .4, size = 1, colour = "#0072B2") +
 geom_smooth(method = "loess", se = FALSE, colour = "#D55E00", linewidth = .8, span = .9) +
 labs(x = "Fitted value", y = "Residual", title = "Residuals vs. fitted")

p_sl <- ggplot(diag_df, aes(fitted, sqrt(abs(resid)))) +
 geom_point(alpha = .4, size = 1, colour = "#0072B2") +
 geom_smooth(method = "loess", se = FALSE, colour = "#D55E00", linewidth = .8, span = .9) +
 labs(x = "Fitted value", y = expression(sqrt(abs(residual))),
 title = "Scale--location (variance check)")

p_rf + p_sl
```

the residual plot below is the required residuals-versus-fitted plot for the preferred specification. The local fit is close to flat with no systematic curvature, which is consistent with a correctly specified conditional mean had the poverty rate been left on its raw scale, the curvature documented above would appear here as a parabolic band.


# Model Building

All specification decisions below are made on the exploration set only. The
confirmation set created in the sampling chunk above is left untouched until
final estimation, so nothing we learn here can leak into the reported results.

We fit three nested specifications. Model 1 regresses test performance on the
poverty rate alone, giving a baseline for the single socioeconomic measure most
often cited in policy discussion. Model 2 adds the remaining five socioeconomic
predictors, which lets us see how much of the poverty association is shared with
income, educational attainment, employment, SNAP receipt, and family structure.
Model 3 adds a squared poverty term to test whether the relationship bends at
higher poverty levels rather than holding to a constant slope.

Because district-level socioeconomic data of this kind routinely shows
non-constant error variance, all standard errors are heteroskedasticity-robust
(HC sandwich estimator) rather than classical.

```{r modeling-setup, message=FALSE, warning=FALSE}
library(lmtest)
library(sandwich)
library(stargazer)

# sample_df and its exploration/confirmation split come from the sampling chunk
# above. stargazer needs a plain data.frame, not a tibble, or it can throw
# "missing value where TRUE/FALSE needed" during table formatting.
explore <- as.data.frame(sample_df %>% filter(split == "exploration"))
cat(sprintf("Exploration set: %d districts\n", nrow(explore)))
```

```{r modeling-fit}
m1 <- lm(cs_mn_avg_ol ~ povertyall, data = explore)

m2 <- lm(cs_mn_avg_ol ~ povertyall + lninc50all + baplusall +
  unempall + snapall + single_momall, data = explore)

m3 <- lm(cs_mn_avg_ol ~ povertyall + I(povertyall^2) + lninc50all +
  baplusall + unempall + snapall + single_momall, data = explore)

# Heteroskedasticity-robust standard errors
robust_se <- function(model) sqrt(diag(vcovHC(model)))
```

```{r modeling-table, results='asis', echo=FALSE}
# stargazer emits its table verbatim (results='asis'), so it bypasses pandoc's
# escaping. A bare "%" in a label would start a LaTeX comment and silently
# swallow the rest of that table row, so escape it for latex output.
pct <- if (isTRUE(knitr::is_latex_output())) "\\%" else "%"

stargazer(m1, m2, m3,
  header = FALSE,
  title = "Regression Results (Exploration Set)",
  se = list(robust_se(m1), robust_se(m2), robust_se(m3)),
  type = if (isTRUE(knitr::is_latex_output())) "latex" else "html",
  dep.var.labels = "Test Performance (SD units from national mean)",
  covariate.labels = c(
    "Poverty Rate", "Poverty Rate Squared", "Log Median Income",
    paste(pct, "Bachelor's Degree+"), "Unemployment Rate",
    paste(pct, "SNAP Recipients"), paste(pct, "Single-Mother Households")
  ),
  digits = 2,
  omit.stat = c("f", "ser"),
  star.cutoffs = c(0.05, 0.01, 0.001)
)
```